On Verification Patterns for Low-Level Systems via Modal Abstractions

### 2025-09-04 paper.tex
+
switch from VMM tech report to OOPSLA citation

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
% Commentout
%\author{Ben Trovato}
%\authornote{Both authors contributed equally to this research.}
%\email{trovato@corporation.com}
%\orcid{1234-5678-9012}
%\author{G.K.M. Tobin}
%\authornotemark[1]
%\email{webmaster@marysville-ohio.com}
%\affiliation{%
 % \institution{Institute for Clarity in Documentation}
 % \city{Dublin}
 % \state{Ohio}
 % \country{USA}
%}

@@

\textit{Contributions} We argue that modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\begin{itemize}
\item \textbf{Identifying System Verification Challenges}: We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\item \textbf{Introducing Modal \emph{Resource Contexts}}: Then we discuss the concept of \emph{resource}, which has already been an essential concept in the design of systems. Inspired by the concept of resource in the systems, we define what a resource and its context are in the modal abstractions.
\item \textbf{Introducing \emph{Nominals for System Resources}}:
Existing logics for systems software must frequently identify
individual resources within a given context ---
nominalization (in the sense of hybrid logic) enables identifying a resource in its context. 
For example, a filesystem transaction is a context of resources for in-memory updated disk blocks
which have not yet been saved to disk.
The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, 
in case of a crash while persisting updated disk-blocks, the filesystem can rollback the 
\emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. 
We discuss other examples to contextualize the existing systems surveyed.
% to be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction 
% identifier -- \emph{strong nominalization}. 
% in another example, virtual memory references (resources) in 
% an address space (resource context), which can be uniquely identified with a root address (the nominal) 
% of its page-table tree, are \emph{agnostic} to the address space that they are in. 
% However, they can only be accessed (be valid) in the address space to which they are agnostic. 
% However, then an address space switch happens, the virtual memory references of the previous address 
% space must be made inaccessible. To be able to do so, although virtual memory references do not hold any 
% piece of information related to their address space, we still have to associate them. 
% We call this kind of unilateral nominalization -- \emph{weak nominalization}.
% \todo[inline,color=cyan]{I left this paragraph alone, mostly, but it's too detailed for the new intro section position here}
\item \textbf{Taxonomy of Current Modal Approaches in System Verification}: Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, reasoning about weak memory models and storage persistence ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2024modalabstractionsvirtualizingmemory,amalreal2024} because we would like to justify that our \emph{definitions} are not domain dependent.
\end{itemize}

@@

\textit{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2024modalabstractionsvirtualizingmemory,amalreal2024}  surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
% \paragraph{Temporal Operators}
% The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
% due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
% distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.

@@

This definition also reveals a subtlety about the denotation of assertions $\phi$ --- they are effectively \emph{predicates} of some kind, functions from some possible world into some logical algebra such as a BI-algebra~\cite{ohearn1999bunched}. 
Taking this interpretation literally gives rise to a convenient way to embed new modalities in \textsf{Iris} 
in a way well-suited to many use cases. 
If $B$ is a type with a BI-algebra structure, then any space of predicate functions (e.g., $\mathsf{val}\rightarrow B$ as a simplification) also carries a BI-algebra structure, lifted from $B$, often called the 
\emph{pointwise lifting} of the algebra. A number of 
projects for systems software verification ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2024modalabstractionsvirtualizingmemory,amalreal2024} now take advantage of this,
explicitly or implicitly lifting the BI-algebra structure from \textsf{Iris} itself to \textsf{Iris} predicates 
used as assertions within modalities. 
Connecting this to \textsf{Iris}'s typeclasses for interactive proofs~\cite{krebbers2018mosel,krebbers-17} then permits working within the embedded logic as in \textsf{Iris}.

@@

\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\scriptsize
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address-Space~\cite{kuru2024modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address-Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

@@

%\paragraph{Weak Context Nominalization} Unlike the resources we discussed so far, some resource elements are \emph{agnostic} to the resource context they belong. That means the resource context needs a nominalization as it needs to be distinguished from other resource contexts when certain actions in the operational semantics are taken (e.g., address-space switch in Table \ref{table:decomposition}), but the resource elements of the context are unaware of this context nominalization, and they may even not be labeled with any other nominal as well. For example, virtual memory references in an address space do not know which address space they belong, but the address spaces are distinguished w.r.t. the unique memory address of the root page-table tree therefore, a virtual-address-points-to ownership assertion does not need to be aware (or nominalized with) of the nominal of the address-space which is the unique memory address of the address table. Logically, specifying an address-space switch ~\cite{kuru2024modalabstractionsvirtualizingmemory} from the address space with the root address $r2$ to the one with the root address $r1$

>>>>

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
% Commentout
\author{Ismail Kuru}
%\authornote{Both authors contributed equally to this research.}
\email{ik335@drexel.edu}
\orcid{0000-0002-5796-2150}
\affiliation{%
  \institution{Drexel University}
  \city{Philadelphia}
  \state{PA}
  \country{USA}
}
\author{Colin Gordon}
%\authornotemark[1]
\email{csg63@drexel.edu}
\orcid{0000-0002-9012-4490}
\affiliation{%
  \institution{Drexel University}
  \city{Philadelphia}
  \state{PA}
  \country{USA}
}

@@

\textit{Contributions} We argue that modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\begin{itemize}
\item \textbf{Identifying System Verification Challenges}: We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\item \textbf{Introducing Modal \emph{Resource Contexts}}: Then we discuss the concept of \emph{resource}, which has already been an essential concept in the design of systems. Inspired by the concept of resource in the systems, we define what a resource and its context are in the modal abstractions.
\item \textbf{Introducing \emph{Nominals for System Resources}}:
Existing logics for systems software must frequently identify
individual resources within a given context ---
nominalization (in the sense of hybrid logic) enables identifying a resource in its context. 
For example, a filesystem transaction is a context of resources for in-memory updated disk blocks
which have not yet been saved to disk.
The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, 
in case of a crash while persisting updated disk-blocks, the filesystem can rollback the 
\emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. 
We discuss other examples to contextualize the existing systems surveyed.
% to be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction 
% identifier -- \emph{strong nominalization}. 
% in another example, virtual memory references (resources) in 
% an address space (resource context), which can be uniquely identified with a root address (the nominal) 
% of its page-table tree, are \emph{agnostic} to the address space that they are in. 
% However, they can only be accessed (be valid) in the address space to which they are agnostic. 
% However, then an address space switch happens, the virtual memory references of the previous address 
% space must be made inaccessible. To be able to do so, although virtual memory references do not hold any 
% piece of information related to their address space, we still have to associate them. 
% We call this kind of unilateral nominalization -- \emph{weak nominalization}.
% \todo[inline,color=cyan]{I left this paragraph alone, mostly, but it's too detailed for the new intro section position here}
\item \textbf{Taxonomy of Current Modal Approaches in System Verification}: Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, reasoning about weak memory models and storage persistence ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024} because we would like to justify that our \emph{definitions} are not domain dependent.
\end{itemize}

@@

\textit{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024}  surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
% \paragraph{Temporal Operators}
% The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
% due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
% distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.

@@

This definition also reveals a subtlety about the denotation of assertions $\phi$ --- they are effectively \emph{predicates} of some kind, functions from some possible world into some logical algebra such as a BI-algebra~\cite{ohearn1999bunched}. 
Taking this interpretation literally gives rise to a convenient way to embed new modalities in \textsf{Iris} 
in a way well-suited to many use cases. 
If $B$ is a type with a BI-algebra structure, then any space of predicate functions (e.g., $\mathsf{val}\rightarrow B$ as a simplification) also carries a BI-algebra structure, lifted from $B$, often called the 
\emph{pointwise lifting} of the algebra. A number of 
projects for systems software verification ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024} now take advantage of this,
explicitly or implicitly lifting the BI-algebra structure from \textsf{Iris} itself to \textsf{Iris} predicates 
used as assertions within modalities. 
Connecting this to \textsf{Iris}'s typeclasses for interactive proofs~\cite{krebbers2018mosel,krebbers-17} then permits working within the embedded logic as in \textsf{Iris}.

@@

\subsection{Decomposing a System into its Constituents \emph{Contingently}}
\label{sec:definitions}
\begin{table*}[t]\scriptsize
\centering
\caption{Modal Decomposition of Program-Logics.}
\label{table:decomposition}
 %$\{\Diamond (P) \}$ write  $\{ P\}_{n}$
%$\{\overset{ICut^{n}}{\hookrightarrow} (P) \}$ return $\{ P\}_{n} $
\begin{tabular}{@{}lcccp{4cm}@{}}
  \toprule
{ Modality}& { Context} & { Elements}  &  { Nominalization} & { Context Steps} \\ \midrule
 Post-Crash~\cite{tejthesis,perennialgit,tejperennial19} &  $\Diamond \; P $  &  $  \ell \mapsto_{n}^{\overline{\gamma}} v $ &  Strong &   Crash Recovery    \\ 
%GC-Modality &      & \multicolumn{2}{c}{\multirow{2}{*}{Because why not}} & 8                                   & & ~\cite{}\\
 NextGen~\cite{larsnextgen25} &   $\overset{t}{\hookrightarrow} \; P$  &  Own (t(a))  &  Strong &  Determined Based on the Model$^{*}$   \\
 StackRegion$^{*}$  ~\cite{larsnextgen25} &  $\overset{ICut^{n}}{\hookrightarrow} \; P$      &  $\fbox{n} \; \ell \mapsto v$                                 &  Strong &  Alloc and Return to/from stack\\
 Memory-Fence~\cite{fsl,fsl++,derekrustbelt20} &  $\triangle_{\pi}$ and $\triangledown_{\pi}$      &  $\ell \mapsto v$                                 &  Weak &  Fence Acquire and Release  \\
 Address-Space~\cite{kuru2025modalabstractionsvirtualizingmemory} &  [r]P     &   $\ell \mapsto v$ &  Weak                         &  Address-Space Switch  \\ 
 Ref-Count~\cite{amalreal2024}& @$_{\ell}$ P    &  $\ell_1 \mapsto v$ &  Weak  &  Allocating, Dropping and Sharing a Reference \\ \bottomrule
\end{tabular}

@@

%\paragraph{Weak Context Nominalization} Unlike the resources we discussed so far, some resource elements are \emph{agnostic} to the resource context they belong. That means the resource context needs a nominalization as it needs to be distinguished from other resource contexts when certain actions in the operational semantics are taken (e.g., address-space switch in Table \ref{table:decomposition}), but the resource elements of the context are unaware of this context nominalization, and they may even not be labeled with any other nominal as well. For example, virtual memory references in an address space do not know which address space they belong, but the address spaces are distinguished w.r.t. the unique memory address of the root page-table tree therefore, a virtual-address-points-to ownership assertion does not need to be aware (or nominalized with) of the nominal of the address-space which is the unique memory address of the address table. Logically, specifying an address-space switch ~\cite{kuru2025modalabstractionsvirtualizingmemory} from the address space with the root address $r2$ to the one with the root address $r1$
